Management induced changes of soil organic carbon on global croplands

### 2020-08-14 SOCbudgetPaper/SOCbudget.tex
update written part to first official version (woohoo)

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

@@

\title{Management induced changes of soil organic carbon on global croplands \newline or \newline How management changed soil organic carbon on global croplands \newline or \newline Agricultural soil have lost 13 GtC topsoil carbon \newline or \newline Linking agricultural management data to soil modeleling -- a global approach}

@@

\begin{abstract}
Soil organic carbon (SOC) is one of larges c stocks on earth (3 times larger biosphere pool).
\end{abstract}

@@

The depletion of soil organic carbon (SOC) due agricultural management has been a major source of carbon emissions, but also offers the potential to turn into a large additional carbon sink and thus mitigate climate change. Whereas the extent of the sink potential by SOC enhancement is highly debated (add refs), it is largely agreed upon that the SOC pool itself is the biggest terrestrial carbon pool, exceeding the atmospheric and even more the biospheric carbon pool multiple times. Even small changes in SOC drivers might lead to substantial shifts in earth carbon cycled and influence the atmospheric CO2 concentration (ref. permafrost melting). The specific amount of carbon stored in the soil is uncertain though and estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile (Bathjes, 1996).

Mapping the worlds SOC has been investigated intensivly over the last decades and led to an increased quality of SOC maps as well as a even better understanding of factors driving magnitude, distribution and dynamics of SOC pools. Natural properties like climatic, biophysical and landscape characteristics clearly play the most important role in this regard.

Human intervention, including land cover change and land management, has however added a further driver to SOC change, which alters terrestrial carbon pools in much shorter time scales and is likely the dominant driver of SOC change on managed land today. Recent studies identified the anthropogenic SOC debt at around 116 GtC (Sanderman et al.), which compares to older estimates of around 60-130 GtC (Lal, 2006). Other studies have focused more closely on spatial disaggregated SOC changes via advanced digital soil mapping techniques (S-World; Stoorvogel 2, 2017) or better representation of biogeochemcial processes within SOC dynamics (). Despite providing a rough estimate of the order of magnitude of change, these studies lack a detailed consideration of land management.

Field-scale models (ref. Daycent, RothC, Ecosse, C-Tool) are able to capture these land mangement impacts by using detailed information on yield levels, fertilizer inputs and various other on farming activities. Yet, due to the lack of comprehensive global management data to feed these input demanding models, it seems still very complex to scale them up to a global extent.

Our study combines an spatial-explicit estimate of agricultural management data on global level with a robust but light weighted SOC model to estimate SOC stocks and stock change factors as well as organic carbon flow dynamics within the agricultural system. We thereby consider change in SOC caused by historical land cover change as well as of different agricultural mangement practices, including residue recycling, manure amendments, irrigation and tillage. We thereby provide the first global, spatial explicit SOC maps that consider agricultural management.

This paper will introduce in the the first part of the method section the basic concept of SOC dynamics. We continue in the second part with a detailed description of the used global gridded management data on production levels, residue recycling rates, manure amendments, irrigation and tillage adoption. In a third, small part, we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines.
In our result section we will focus on the SOC dynamics on global croplands by analysing (1) the spatial explicit SOC distribution and depletion over the last decades, (2) compare the climate zone specfic stock changes with default stock changes factors from the IPCC Tier 1 approach and (3) compare global agricultrual carbon flows and stocks to contextualize the importants of the different management aspects.
Finally, we will discuss our findings and their implications on SOC model development and conlude with an analysis and outlook on the the opportunities for SOC management for climate change mitigation and negative emissions.

\hypertarget{method-50}{%
\section{Method (50)}\label{method-50}}

We compiled calculations as open-source R packages available under github.com/pik-piam/mrcommons (management related functions) and github.com/pik-piam/mrSOCbudget (soil dynamic related functions), which are both based on github.com/pik-piam/madrat, a framework improving reproducibility and transparency in data processing.
In the following chapter we outine the most important relationships and assumptions. See table \ref{append:subsection2mrfunctions} for further information on corresponding code within the R packages.

\hypertarget{sec:carbonbudget}{%
\subsection{Carbon Stocks following (new) Tier 2 method (50)}\label{sec:carbonbudget}}

Following the tier 2 approach of the refinement of IPCC guidelines vol.~4 (\citet{ipcc_2019_2019}), we estimate soil organic carbon (SOC) stocks for cropland and all other land represented by natural vegetation on half-degree resolution from 1975 to 2010. We assume the actual SOC state converges towards a stable steady state, that itself is changing over time and space depending on biophysical, climatic and agronomic conditions. Therefore we conduct the following three steps within each yearly timestep:
(1) We calculate annual land-use (sub-)type specific steady states and decay rates for SOC stocks,
(2) We account for land conversion by transferring SOC between land-use types and
(3) We estimate SOC stocks based on the stocks of the previous time period, the steady state stocks and the decay rate.

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks \(SOC^{eq}\) are given by
\begin{equation}
SOC^{eq} =\frac{C^{\textrm{in}}}{k} \qquad\forall\quad i,t
\label{eq:inoutflow}
\end{equation}
with \(C^{\textrm{in}}\) being carbon inputs to the soil, \(k\) denoting the soil organic carbon decay rate; as well as \(i\) representing grid cells indecies and \(t\) years. We use for our calculations the steady-state method of the refinement of the IPCC guidelines vol.~4 (\citet{ipcc_2019_2019}) for mineral soils, which assume three soil carbon sub-pools (active, slow and passive) and entangled dynamics between them. Annual carbon inflow to each sub-pool (see \ref{sec:carboninputs}) and annual decay rates (see \ref{sec:tier2}) of each sub-pool are still the key components to determining steady-state SOC stocks.

\hypertarget{carbon-inputs-to-the-soil-seccarboninputs}{%
\paragraph*{Carbon Inputs to the Soil \{\#sec:carboninputs\}}\label{carbon-inputs-to-the-soil-seccarboninputs}}
\addcontentsline{toc}{paragraph}{Carbon Inputs to the Soil \{\#sec:carboninputs\}}

We account for different carbon input sources depending on our two distinguished land-use types (see table \ref{tab:datasourceinputs}); in particular for recycled crop residues, below ground biomass of crops (for both see \ref{sec:residues}) and recycled manure (see @ref(sec:livst\_manure)) on croplands; and litterfall (\citep{LPJmL4_1}) on natural vegetation.

Following the IPCC methodology carbon inputs are disaggregated into metabolic and structural components depending on their lignin and nitrogen content (see \citep{ipcc_2019_2019}). For each component the sum over all carbon input sources is allocated to the respective SOC sub-pools via transfer coefficients. This implies that not only the amount of carbon, but also their structural composition is determining the effective inflow. Data sources for all considered carbon inputs as well as for lignin and nitrogen content can be found in table \ref{tab:datasourceinputs}.

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land-use types }
 \begin{tabular}{l l l l}
 \tophline
  \textbf{land-use types}   & \textbf{source of carbon inputs} & \textbf{data source} & \textbf{nitrogen and lignin content} \\
 \middlehline
 \multirow{3}{*}{Cropland} & residues & FAOSTAT, LPJmL4 [2, \\ref{sec:residues}] & default values given by [2]  \\
                            & dead below ground biomass of crops & FAOSTAT, LPJmL4 [2, \\ref{sec:residues}] & default values given by [2] \\
                            & manure & FAOSTAT, Isabelle [2, sec:manure] & default values given by [2] \\
                            \hline
  Natural vegetation        & annual litterfall & LPJmL4 [4]& \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut Nitrogen and lignin content of tree compartments used in CENTURY [4] \strut \end{minipage}\tabularnewline
 \bottomhline
 \end{tabular}
 \belowtable{}
 \label{tab:datasourceinputs}
 \end{table*}

\hypertarget{soil-organic-carbon-decay-300-sectier2}{%
\paragraph*{Soil Organic Carbon decay (300) \{\#sec:tier2\}}\label{soil-organic-carbon-decay-300-sectier2}}
\addcontentsline{toc}{paragraph}{Soil Organic Carbon decay (300) \{\#sec:tier2\}}

The sub-pool specific decay rates are influenced by climatic conditions, biophysical and biochemical soil properties as well as management factors that all vary over time (t) and space (i). Following the steady-state method of the refinement of the IPCC guidelines vol.~4 (\citep{ipcc_2019_2019}) for mineral soils we consider temperature (temp), water (wat), sand fraction (sf) and tillage (till) effects to account for spatial variation of decay rates. Thus \(k_{sub}\) is given by

@@

For cropland we distinguish the effect of different tillage (see \ref{sec:tillage}) and irrigation (see \ref{sec:irrigation}) practices on decay rates, whereas on natural vegetation, we assume rainfed and non-tilled conditions. Data sources as well as considered effects for each land-use types are shown in table \ref{tab:datasourcedecay}. To account for variations of decay rates within each grid cell due to different tillage and irrigation regimes, average rates based on area shares are calculated.

@@

\hypertarget{total-soc-stocks}{%
\subsubsection{Total SOC stocks}\label{total-soc-stocks}}

Carbon stocks \(SOC\) for each sub-pool (sub) converge towards the calculated steady-state stock \(SOC^{eq}\) for each land-use types (lut), each sub-pool (sub) and each annual time step (t) as represented in equation \eqref{eq:steadystate}.

@@

Reformulated equation \eqref{eq:SOCstate}, we obtain a massbalance equation as follows

\begin{equation}
SOC_{t} = SOC_{t-1} - \underbrace{SOC_{t-1} \cdot k_{t} \cdot 1\unit{a}}_{\text{outflow}} + \overbrace{SOC^{eq}_{t} \cdot k_{t} \cdot 1\unit{a}}^{\text{input (using equation \\eqref{eq:inoutflow})}}  \qquad \forall\quad lut, sub, i.
\label{eq:steadystate2budget}
\end{equation}

@@

\begin{equation}
SOC_{t} = \sum_{i} \underbrace{\sum_{lut} \overbrace{\sum_{sub} SOC_{lut, sub, t, i}.}^{\text{$SOC_{lut, t, i}$ - land-use type specific SOC stock within cell}}}_{\text{$SOC_{t, i}$ - total SOC stock within cell}}
\label{eq:totalstock}
\end{equation}

@@

To initialize all SOC sub-pools we assume that cropped land as well as natural vegetation are in a steady state for the specific configuration present within the initilzation year 1961. We assume after a spin-up period of 15 years reliable results start from 1975 and improve over time, since dependency for the initial conditions will decrease.

@@

\hypertarget{sec:agrimanagement}{%
\subsection{Agricultural management data on 0.5 degrees (50)}\label{sec:agrimanagement}}

We compile country-specific FAO production and cropland statistics (\citep{FAOSTAT}) to a comprehensive and constistent data suite. The data is prepared in 5 year time steps from 1965 to 2010, which also restricts our analysis together with the accounting for a spin up phase to time span from 1975 to 2010. For all the following data, if not declared differently, we interpolate values linearly between the time steps.

\hypertarget{sec:landuse}{%
\subsubsection{Landuse and Landuse Change (150)}\label{sec:landuse}}

Land-use patterns are based on the Land-Use Harmonization 2 (LUH2, \citep{LUH2}) data set, which we aggregate from quarter degree to half degree resolution. We disaggregate the five different cropland subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17 crop groups, applying the relative shares for each gridcell based on the country and year specific area harvested shares of FAOSTAT data (\citep{FAOSTAT}) (see @ref(append:Table\_luh2fao2mag) for more details on the crop group mapping). Land-use transitions are calculated as net area differences of the land-use data on half-degree.

\hypertarget{sec:residues}{%
\subsubsection{Crop and Crop Residues (300)}\label{sec:residues}}

\textbf{Crop Production}
Using half-degree yield data from LPJmL (\citep{LPJmL4_1}) as well as half-degree cropland patterns (see \ref{sec:landuse}) we compile crop group specific half-degree production patterns. We calibrate cellular yields with one country-level calibration factor for each crop group to meet historical FAOSTAT production (\citep{FAOSTAT}). Note that by using physical cropland areas we account for multiple crop harvest events as well as for fallows.

\textbf{Crop Residue Production}
Crop residue production and management is based on a revised methodology of (\citep{bodirsky2012}) and will be explained in key aspects again due to its central role for soil carbon modelling. Starting from crop production (\(CP\)) estimates of the harvested organs and their respective harvested crop area (\(CA\)), we estimate above-ground (\(AGR\)) and below-ground (\(BGR\)) residual biomass using crop group (\(cg\)) specific harvest indices (\(HI\)) and root:shoot ratios (\(RS\)) as follows

@@

Following the IPCC guidelines, we split the harvest index into a yield and an area dependend fraction (\citep{ipcc_2006_2006}). Note that deviating from \citep{bodirsky2012} we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased amounts on fallow land.
We assume that all BGR are recycled to the soil, whereas AGR can be burned or harvested for other purposes such as feeding animals (\citep{weindl}), fuel or for material use.

\textbf{Burned Residues}
A fixed share of the AGR is assumed to be burned on field depending on the per-capita income of the country. Following \citep{smil1999}) we assume 25\% burn share for low-income countries according to worldbank definitions (\(<\,1000\,\tfrac{USD}{yr}\)), 15\% for high-income (\(>\,10000\,\tfrac{USD}{yr}\) and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop group 80--90\% of the residue carbon burned on the fields are lost within the combustion process (\citep{ipcc_2006_2006}).

\textbf{Residue Usage}
We compile out of our 17 crop groups, three used residue groups (straw, high-lignin and low-lignin residues) with additional demand for other purposes and one residues with no double use (see @ref(append:Table\_kcr2kres)). Residue feed demand for five different livestock groups is based on country- and residue-group-specific feed basekts (see \citep{weindl}) taking available AGR biomass as well as livestock productivity into account. We estimate a material-use share for the straw residues group of 5\% and a fuel-share of 10\% for all used residues groups in low income countries according to worldbank definitions (\(<\,1000\,\tfrac{USD}{yr}\)). For high-income (\(>\,10000\,\tfrac{USD}{yr}\) no withdrawl for material or fuel use is assumend, leaving middle-income countries with linearly interpolate shares depending on their per-capita income. The remaining AGR as well as all BGR are assumend to be recycled to the soil. We cut high recycling shares at \(10\tfrac{\unit{tC}}{\unit{ha}}\) to corrected for outliers.

\textbf{Dry Matter to Carbon Transformation}
To transform dry matter estimates into carbon, we compiled crop-group and plant part specific carbon to dry matter (c:dm) ratios (see @ref(append:Table\_c2dm)) (\citet{Ma2018}).

\hypertarget{sec:livst_manure}{%
\subsubsection{Livestock Distribution and Manure Excretion (300)}\label{sec:livst_manure}}

@@

\hypertarget{sec:tier1}{%
\subsection{Carbon Budget following Tier 1 (150)}\label{sec:tier1}}

>>>>

@@

\title{Management induced changes of soil organic carbon on global croplands}

@@

\begin{abstract}
Soil organic carbon (SOC) is one of larges carbon stocks on earth. It is three times larger than the biospheric pool and more than twice as big as the athmospheric pool, when looking into the first meter of the earth soil profile. Human cropping activties led and still lead to a depletion of SOC though, which are so far not well represented in global assessments of historic carbon emissions. While SOC models often represent well the biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated on the global scale. Here we create a spatial explicit data set for agricutural management on cropland, especially for crop residue and manure management, based on global historic production (FAOSTAT), and land-use (LUH2) data; and combine it with the IPCC Tier 2 approach to create a half degree resolution SOC stock changes on mineral soils. We estimate that due to arable farming soils have lost over 37 GtOC of which 4 GtOC have been regained within the period 1975-2010. We show that, our results on global scale based on Tier 2 IPCC methodolgy are in good agreement with Tier 1 default assumptions. We also find that SOC is very sensitive to management decision such as residue recycling indicating the nessessity to incorporated better management data in soil models.
\end{abstract}

@@

Possible titles:

\begin{itemize}
\item
  How management changed soil organic carbon on global croplands
\item
  Agricultural soil have lost 37 GtC topsoil carbon
\item
  Linking agricultural management data to soil modeleling -- a global approach
\end{itemize}

Soil Organic Carbon (SOC), the amount of organic carbon stored through the earth's soil, is the largest terrestrial carbon pool, exceeding the carbon in the atmospheric and biospheric pools multiple times (Bathjes). As such, even small changes in drivers of SOC may thus lead to substantial shifts in earth carbon cycle and influence the amount of CO2 in the atmosphere (ref. permafrost melting). The specific amount of carbon stored in the soil is however uncertain, with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile (Bathjes, 1996). The quality of SOC maps has markedly improved in recent decades, along with better understanding of the factors driving the global magnitude, distribution, and dynamics of SOC pools. Natural properties like climatic, biophysical, and landscape characteristics clearly play the most important roles in this regard.\\
Human intervention, including land cover change and land management, has however added a further driver to SOC change, which alters terrestrial carbon pools in much shorter time scales and is likely one of the most dominant driver of SOC changes on managed land today. Recent studies identify the anthropogenic SOC debt for the first meter of the soil profile at around 116 GtC (Sanderman et al.), compared to previous estimates between 60-130 GtC (Lal, 2006). Other studies have focused more closely on spatially disaggregation of SOC changes via advanced digital soil mapping techniques (S-World; Stoorvogel 2, 2017) or better representation of biogeochemical processes within SOC dynamics (Hararuk, 20XX). While providing a rough estimate of the order of magnitude of change, these studies lack a detailed consideration of land management, especial agricultural activities.\\
Field-scale models (ref. Daycent, RothC, Ecosse, C-Tool) are able to capture these land mangement impacts by using detailed information on crop yield levels, fertilizer inputs and various other on-farm activities. However, due to the lack of comprehensive global management data as input to these models, scaling up to the global extent remains a complex challenge.\\
Our study combines an spatially explicit estimate of agricultural management data on the global level with a 3-pool SOC model parametrized for global croplands. This allows us to estimate SOC stock change factors, as well as organic carbon flow dynamics within the agricultural system. We thereby consider change in SOC caused by historical land cover change as well as of different agricultural mangement practices, including residue recycling, manure amendments, irrigation, and tillage. We thereby provide the first global, spatially explicit SOC loss maps that consider detailed agricultural management.\\
This paper will first introduce in the methods section the basic concept of SOC dynamics as applied in this study. We continue with a detailed description of the global gridded management data used here, including crop production levels, residue recycling rates, manure amendments, and the adoption of irrigation and tillage practices. Lastly, we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines. In our results section we focus on the SOC dynamics of global croplands by (1) analysing the spatially explicit distribution and depletion of SOC from 1975 to 2010, (2) comparing the climate zone specfic stock changes with default stock change factors from the IPCC Tier 1 approach and (3) quantify global agricultural carbon flows and stocks to compare the importance of various management aspects. Finally, we will discuss our findings -- including their implications for SOC model development -- and conclude with an outlook on the ability of SOC management to mitigate climate change and contribute to negative greenhouse gas emissions.\\
\newpage

\hypertarget{method}{%
\section{Method}\label{method}}

We compile calculations as open-source R packages available at github.com/pik-piam/mrcommons (management related functions) and github.com/pik-piam/mrSOCbudget (soil dynamic related functions), which are both based on the MADRaT package (package citation?), a framework which aims to improve reproducibility and transparency in data processing.
In the following chapter we outine the most important relationships and assumptions. Table \ref{append:subsection2mrfunctions} provides further information on corresponding code within the R packages.

\hypertarget{sec:carbonbudget}{%
\subsection{SOC Stocks and Stock changes following the Tier 2 steady-state method}\label{sec:carbonbudget}}

Following the tier 2 approach of the refinement of IPCC guidelines vol.~4 (\citet{ipcc_2019_2019}; short Tier 2 steady-state method) , we estimate soil organic carbon (SOC) stocks and their change over time for cropland on half-degree resolution from 1975 to 2010. We assume the current SOC state converges towards a steady state, which itself is depending on biophysical, climatic and agronomic conditions. Therefore we conduct the following three steps within each yearly timestep:
(1) We calculate annual land-use (sub-)type-specific steady states and decay rates for SOC stocks,
(2) we account for land conversion by transferring SOC from and to an other land-use type representad as natural vegetation,
(3) we estimate SOC stocks and changes based on the stocks of the previous time period, the steady state stocks and the decay rate.

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks \(SOC^{eq}\) are given by
\begin{equation}
SOC^{eq} =\frac{C^{\textrm{in}}}{k} \qquad\forall\quad i,t
\label{eq:inoutflow}
\end{equation}
with \(C^{\textrm{in}}\) being carbon inputs to the soil, \(k\) denoting the soil organic carbon decay rate; as well as \(i\) representing grid cell indices and \(t\) years. We use for our calculations the Tier 2 steady-state method, which assumes three soil carbon sub-pools (active, slow and passive) and interactions between them. Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are the key components to determining steady-state SOC stocks.

\textbf{Carbon Inputs to the Soil}

We account for different carbon input sources depending on the two land-use types we distinguish: croplands and natural vegetated land as representative for all other land use (see table \ref{tab:datasourceinputs}). Carbon sources for cropland are recycled crop residues, below ground biomass of crops (for both see \ref{sec:residues}) and recycled manure (see \ref{sec:livstmanure}); for natural vegetation litterfall \citep{schaphoff_lpjml4_2018} is the only source of carbon inflow to the soil.

Following the IPCC carbon accounting methodology, carbon inputs are disaggregated into metabolic and structural components depending on their lignin and nitrogen content (see \citet{ipcc_2019_2019}). For each component the sum of all carbon input sources is allocated to the respective SOC sub-pools via transfer coefficients. This implies that both the amount of carbon as well as its structural composition determine the effective inflow. Data sources for all considered carbon inputs as well as for lignin and nitrogen content can be found in table \ref{tab:datasourceinputs}.

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land-use types }
 \begin{tabular}{l l l l}
 \tophline
  \textbf{land-use types}   & \textbf{source of carbon inputs} & \textbf{data source} & \textbf{nitrogen and lignin content} \\
 \middlehline
 \multirow{3}{*}{Cropland} & residues & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2]  \\
                            & dead below ground biomass of crops & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2] \\
                            & manure & FAOSTAT, Isabelle [2, sec:manure] & default values given by [2] \\
                            \hline
  Natural vegetation        & annual litterfall & LPJmL4 [4]& \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut Nitrogen and lignin content of tree compartments used in CENTURY [4] \strut \end{minipage}\tabularnewline
 \bottomhline
 \end{tabular}
 \label{tab:datasourceinputs}
 \belowtable{}
 \end{table*}

\textbf{SOC decay}

The sub-pool specific decay rates are influenced by climatic conditions, biophysical and biochemical soil properties as well as management factors that all vary over time (t) and space (i). Following the Tier 2 steady-state method we consider temperature (temp), water (wat), sand fraction (sf) and tillage (till) effects to account for spatial variation of decay rates. Thus \(k_{sub}\) is given by

@@

For cropland we distinguish the effect of different tillage (see \ref{sec:tillage}) and irrigation (see \ref{sec:irrigation}) practices on decay rates, whereas for natural vegetation, we assume rainfed and non-tilled conditions. Data sources as well as drivers considered for each land-use types are shown in table \ref{tab:datasourcedecay}. To account for variations of decay rates within each grid cell due to different tillage and irrigation regimes, average rates based on area shares are calculated.

@@

\hypertarget{total-soc-stocks-and-stock-changes}{%
\subsubsection{Total SOC stocks and stock changes}\label{total-soc-stocks-and-stock-changes}}

Carbon stocks \(SOC\) for each sub-pool (sub) converge towards the calculated steady-state stock \(SOC^{eq}\) for each land-use type (lut), each sub-pool (sub) and each annual time step (t) like

@@

Reformulating this equation, we obtain a massbalance equation as follows

\begin{equation}
SOC_{t} = SOC_{t-1} - \underbrace{SOC_{t-1} \cdot k_{t} \cdot 1\unit{a}}_{\text{outflow}} + \overbrace{SOC^{eq}_{t} \cdot k_{t} \cdot 1\unit{a}}^{\text{input (using equation (1))}}  \qquad \forall\quad lut, sub, i.
\label{eq:steadystate2budget}
\end{equation}

@@

\begin{equation}
SOC_{t} = \sum_{i} \underbrace{\sum_{lut} \overbrace{\sum_{sub} SOC_{lut, sub, t, i}.}^{\text{$SOC_{lut, t, i}$ -- land-use type specific SOC stock within cell}}}_{\text{$SOC_{t, i}$ -- total SOC stock within cell}}
\label{eq:totalstock}
\end{equation}

According to the IPCC guidelines SOC changes can be calculates as the difference of two (aufeinander folgenden) years (see Eq. 5.0A in \citep{ipcc_2019_2019}). This however will also include naturally occured changes due to climatic variance over time. For our study we will define the absolute and relative SOC changes in relation to a potential natural state \(SOC_{pnv}\) under the same climatic conditions at time \(t\) in grid cell \(i\), that is based on the natural vegetation SOC calculations as defined above without accounting for land conversion from cropland at any time. The absolute changes \(\Delta SOC\) and relative changes \(F^{SCF}\) a thus given by

\begin{equation}
\Delta SOC_{t, i} = SOC_{t, i} - SOC_{pnv, t, i}\quad, \qquad \quad  F^{SCF}_{t, i} = \frac{SOC_{t, i}}{SOC_{pnv, t, i}}.
\label{eq:totalstock}
\end{equation}

Note that the absolution changes \(\Delta SOC\) can be also interpreted as cummulative SOC emissions or as SOC depth of mankind; whereas relative changes \(F^{SCF}\) can be denoted as stock change factors as defined within the IPCC guidelines of 2006 \citep{ipcc_2006_2006}.

@@

To initialize all SOC sub-pools we assume that cropped land as well as natural vegetation are in a steady state for the specific configuration present within the initialization year 1961. We assume after a spin-up period of 15 years reliable results start from 1975 and improve over time, since dependency on the initial conditions will decrease.

@@

\hypertarget{sec:agrimanagement}{%
\subsection{Agricultural management data on 0.5 degree resolution}\label{sec:agrimanagement}}

We compile country-specific FAO production and cropland statistics \citep{faostat_faostat_2016} to a harmonized and consistent data set. The data is prepared in 5 year time steps from 1965 to 2010, which, together with the spin up phase, restricts our analysis to time span from 1975 to 2010. For all the following data, if not declared differently, we interpolate values linearly between the time steps and hold them constant before 1965.

\hypertarget{sec:landuse}{%
\subsubsection{Landuse and Landuse Change}\label{sec:landuse}}

Land-use patterns are based on the Land-Use Harmonization 2 \citep{Hurtt2020} data set, which we aggregate from quarter degree to half degree resolution. We disaggregate the five different cropland subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17 crop groups, applying the relative shares for each grid cell based on the country and year specific area harvested shares of FAOSTAT data \citep{faostat_faostat_2016} (see \ref{append:Tableluh2fao2mag} for more details on the crop group mapping). Land-use transitions are calculated as net area differences of the land-use data on half-degree.

\hypertarget{sec:residues}{%
\subsubsection{Crop and Crop Residues Production}\label{sec:residues}}

\textbf{Crop Production}

Using half-degree yield data from LPJmL \citep{schaphoff_lpjml4_2018} as well as half-degree cropland patterns (see \ref{sec:landuse}) we compile crop group specific half-degree production patterns. We calibrate cellular yields with a country-level calibration factor for each crop group to meet historical FAOSTAT production \citep{faostat_faostat_2016}. Note that by using physical cropland areas we account for multiple crop harvest events as well as for fallows.

\textbf{Crop Residue Production}

Crop residue production and management is based on a revised methodology of \citep{bodirsky_n2o_2012} and key aspects are explained here given its central role for soil carbon modelling. Starting from harvested crop production (\(CP\)) estimates and their respective harvested crop area (\(CA\)), we estimate above-ground (\(AGR\)) and below-ground (\(BGR\)) residual biomass using crop group (\(cg\)) specific harvest indices (\(HI\)) and root:shoot ratios (\(RS\)) as follows

@@

Following the IPCC guidelines, we split the harvest index into a yield and an area dependend fraction \citep{ipcc_2006_2006}. Note that deviating from \citep{bodirsky_n2o_2012} we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased amounts on fallow land.
We assume that all BGR are recycled to the soil, whereas AGR can be burned or harvested for other purposes such as feeding animals \citep{weindl}, fuel or for material use.

\textbf{Burned Residues}

A fixed share of the AGR is assumed to be burned on field depending on the per-capita income of the country. Following \citep{smil1999}) we assume 25\% burn share for low-income countries according to worldbank definitions (\(<\,1000\,\tfrac{USD}{yr\,cap}\)), 15\% for high-income (\(>\,10000\,\tfrac{USD}{yr\,cap}\)) and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop group 80--90\% of the residue carbon burned on the fields are lost within the combustion process \citep{ipcc_2006_2006}.

\textbf{Residue Usage}

We compile out of our 17 crop groups, three residue groups (straw, high-lignin and low-lignin residues) with additional demand for other purposes and one residue with no double use (see \ref{append:Tablekcr2kres}). Residue feed demand for five different livestock groups is based on country- and residue-group-specific feed baskets (see \citep{weindl}), taking available AGR biomass as well as livestock productivity into account. We estimate, for low-income countries, a material-use share for straw residues of 5\% and a fuel-share of 10\% for all used residues groups in low-income countries. For high-income countries, no withdrawal for material or fuel use is assumend, and middle-income countries use shares are linearly interpolated based on per-capita income. The remaining AGR as well as all BGR are assumend to be recycled to the soil. We limit high recycling shares at \(10\tfrac{\unit{tC}}{\unit{ha}}\) to correct for outliers.

\textbf{Dry Matter to Carbon Transformation}

To transform dry matter estimates into carbon, we compiled crop-group and plant part specific carbon to dry matter (c:dm) ratios (see \ref{append:Tablec2dm}).

\hypertarget{sec:livstmanure}{%
\subsubsection{Livestock Distribution and Manure Excretion}\label{sec:livstmanure}}

@@

\hypertarget{sec:tier1}{%
\subsection{SOC Stocks and Stock changes following Tier 1}\label{sec:tier1}}

Additionally to the tier 2 approach of the refinement of IPCC guidelines vol.~4 \citep{ipcc_2019_2019} and the detailed analysis of management data coming with it, SOC changes can be estimated using the IPCC tier 1 approach of IPCC guidelines vol.~4 (\citet{ipcc_2006_2006}, \citet{ipcc_2019_2019}). Here, stocks are calculated via stock change factors (F\^{}\{SCF\}) given by the IPCC for the topsoil (0-30 cm) and based on observational data. \(F^{SCF}\) are differentiated by different crop, management and input systems (here summarized under \(m\)) reflecting different dynamics under changed in- and outflows without explicitly tracking these. Moreover \(F^{SCF}\) vary for different climatic zones (\(c\)) specified by the IPCC (see \ref{append:climatemap}). The actual SOC stocks as thus calulated based on a given reference SOC stock by

\begin{equation}
SOC^{\text{target}}_{i} = \sum_{c,m} T_{c,i} \cdot SOC^{\text{ref}} \cdot F^{SCF}_{c,m} \qquad\forall\quad t,
\label{eq:tier1}
\end{equation}
